---
title: "Week 1: Finding Census Data"
author: "Carole Voulgaris"
date: "9/3/2020"
output: 
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 3
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

The tidycensus package offers a really convenient way to download data collected by the United States Census Bureau directly into your R environment. The most recent update to tidycensus allows you to access microdata (e.g. data at the person and household levels) in addition to area-level data (census tracts, counties, cities, etc). 

You can read more about tidycensus here: https://walker-data.com/tidycensus/

We'll be using a pre-release (development) version of tidycensus, so you'll need to install it by running the following code in your console (no need to save it in your Rmarkdown file since you'll only need to ever do this once):

`remotes::install_github("walkerke/tidycensus")`

Once you've done that, load the following libraries:

```{r load libraries}
library(tidycensus)
library(tidyverse)
```

You will also need to obtain a census api key from http://api.census.gov/data/key_signup.html and install it on your computer by typing `census_api_key("YOUR API KEY GOES HERE", install=TRUE)` into your console. You only need to install libraries and your API key once, but you'll need to load the libraries every time you run your code.

## Finding variables
Before we can download the data, we need to decide which variables we want, so we need to see what variables we have to choose from.

Let's start with microdata. What variables are available if I want do do a person-level analysis of data from the 1-year American Community Survey Sample from 2018?

```{r load person-level microdata variables}
person_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code, 
           var_label, data_type, level) %>%
  filter(level == "person", year == 2018, survey == "acs1")
```

Now I can look through the person_vars_2018 dataframe in my viewer to browse variables.

And what if I want household-level variables from the same survey?

```{r load household-level microdata variables}
hh_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code, 
           var_label, data_type, level) %>%
  filter(level == "housing", year == 2018, survey == "acs1")
```

I can also get data at an area level. The area-level variables will be the same for any geography (although they'll differ between the decennial census and the American Community survey). 

What area-level variables are available from the American Community Survey for the 2018 1-year sample?

``` {r load ACS area variables}
area_vars_2018 <- load_variables(2018, "acs1")
```

What area-level variables are available from the 2010 census?

```{r load decennial variables}
area_vars_2010 <- load_variables(2010, "sf1")
```

## Loading data

Let's say I've decided to to a person-level analysis. After looking through the variables available in `person_vars_2018`, I've chosen the following five variables: 

1. Housing tenure (TEN) (categorical)
2. Sex (SEX) (categorical)
3. Income (PINCP) (continuous)
4. Work hours per week (WKHP) (continuous)
5. Travel time to work (JWMNP) (continuous)

And I want my sample to include all survey respondents in Texas who travel to work (everyone with a value greater than zero for travel time to work).

Note that since every person is in a household, it's fine to include household-level variables (like tenure) in a person-level analysis

``` {r load person-level data, results = 'hide'}
person_data <- get_pums(variables = c("TEN", 
                                      "SEX", 
                                      "PINCP", 
                                      "WKHP", 
                                      "JWMNP"),
                        state = "TX", 
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
  filter(JWMNP > 0) %>%
  select(TEN_label, SEX_label, PINCP, WKHP, JWMNP)
```

You can type the name of the dataframe you created to show the first few rows.
``` {r show person-level data}
person_data
```

The method for getting household-level data is the same. You just need to make sure all of your variables are household-level variables. I'll get this set of variables for all rented homes in New York state.

1. Age of structure (YBL) (categories we'll convert to continuous)
2. Number of bedrooms (BDSP) (continuous)
3. Monthly rent (GRNTP) (continuous)
4. Units in structure (BLD) (categorical)
5. Limited English-speaking household (LNGI) (categorical)

``` {r load household-level data, results = 'hide'}
hh_data <- get_pums(variables = c("TEN", 
                                  "YBL", 
                                  "BDSP", 
                                  "GRNTP", 
                                  "BLD",
                                  "LNGI"),
                        state = "NY", 
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
  filter(TEN_label == "Rented") %>%
  mutate(age_struct = case_when(
    YBL_label == "1939 or earlier" ~ 100,
    YBL_label == "1940 to 1949" ~ 73,
    YBL_label == "1950 to 1959" ~ 63,
    YBL_label == "1960 to 1969" ~ 53,
    YBL_label == "1970 to 1979" ~ 43,
    YBL_label == "1980 to 1989" ~ 33,
    YBL_label == "1990 to 1999" ~ 23,
    YBL_label == "2000 to 2004" ~ 16,
    YBL_label == "2005" ~ 13,
    YBL_label == "2006" ~ 12,
    YBL_label == "2007" ~ 11,
    YBL_label == "2008" ~ 10,
    YBL_label == "2009" ~ 9,
    YBL_label == "2010" ~ 8,
    YBL_label == "2011" ~ 7,
    YBL_label == "2012" ~ 6,
    YBL_label == "2013" ~ 5,
    YBL_label == "2014" ~ 4,
    YBL_label == "2015" ~ 3,
    YBL_label == "2016" ~ 2,
    YBL_label == "2017" ~ 1,
    YBL_label == "2018" ~ 0)) %>%
  select(age_struct, BDSP, GRNTP, BLD_label, LNGI_label)
```

```{r show household data}

hh_data
  
```

Or maybe I want to do an area-level analysis. I'll get (or calculate) the following five variables for each county in the United States:

1. Percent of population with a college degree (age 25 to 64) (continuous)
2. Unemployment rate (age 16 and over) (continuous)
3. Median income (continuous)
4. Racial composition (majority white or majority minority) (categorical)
5. Rural (100% rural population from 2010 census) (categorical)

``` {r load county-level data}
pop_vars <- c(workers_16 = "B23025_002",
          unemp_pop_16 = "B23025_005",
          college_25 = "B23006_023",
          popln_25 = "B23006_001",
          med_income = "B06011_001",
          tot_pop = "B01003_001",
          white_pop = "B01001A_001")

rural_vars <- c(total_hh = "H002001",
                rural_hh = "H002005")

counties2018 <- get_acs(geography = "county", 
                        variables = pop_vars, 
                        output = "wide",
                        year  = 2018,
                        survey = "acs1")

counties_rural <- get_decennial(geography = "county", 
                                variables = rural_vars, 
                                output = "wide",
                                year = 2010)

counties <- inner_join(counties2018, counties_rural, by = "GEOID") %>%
  mutate(pct_college = college_25E / popln_25E,
         unemp_rate = unemp_pop_16E / workers_16E,
         majority_white = white_popE / tot_popE > 0.5,
         rural = rural_hh == total_hh) %>%
  select(pct_college, unemp_rate, med_incomeE, 
                majority_white, rural)

counties
```

## Saving data

You can save any of the datasets you just created for future use by writing them to a csv file. 

``` {r save dataframes as csv}
write_csv(person_data, "people.csv")
write_csv(hh_data, "households.csv")
write_csv(counties, "counties.csv")
```